Developing Observational Health Research Capacity in the Julia Ecosystem

### 2023-12-12 paper.tex
Continuing massive reviews of the JuliaHealth paper

% EXPLAIN: Why is this a crucial innovation?
% EXPLAIN: What is meant by a transferable analysis
Arguably, the most crucial innovation that has emerged is the Observational Medical Outcomes Partnership Common Data Model (OMOP CDM) which was designed specifically to standardize observational health data to enable transferable analyses.
% TODO: Add citation
% NOTE: Why is mapping relevant here?
The importance of this model has become apparent within open science organizations like the OHDSI (Observational Health Data Science and Informatics) network where members across $100+$ countries with access to various health systems have been able to successfully map their patient data to the OMOP CDM.
% TODO: Add citation about successful collaborations
% EXPLAIN: What is a network study
% TODO: Cite the quote "write once, run everywhere"
This has led to successful collaboration relationships across countries where one research group can develop a network study and then disseminate the study across collaborators leading to the idea that you can "write once, run everywhere."

@@

JuliaHealth is an open-science collective dedicated to bringing the Julia programming language to various avenues of health research domains including, but not limited to, observational health, medical imaging, standards interoperability and more.
It leverages the strengths of the Julia language to introduce new methods into these spaces, augment existing workflows, and provide a community structure for health researchers within the global Julia ecosystem.
Eventually, JuliaHealth will host a variety of subecosystems and associated resources that will enable novel research to be conducted within Julia.

As JuliaHealth is an ad hoc network of contributors, varying aspects of the ecosystem have developed at differing rates as a function of factors such as developer time or workplace incentives.
One area of the subecosystems within JuliaHealth that has matured significantly over the years has been the JuliaHealth observational health subecosystem (referred to as OHS or the "subecosystem" for the remainder of this paper).

Although difficult to date, it can said that OHS had its inception in work done by Evans and Simonov with efforts such as \textit{FunOHDSI.jl}, \textit{OHDSICohortExpressions.jl}, and \textit{FunSQL.jl} around $2021$ \cite{kirill_simonov_2023_7705325}
Since then, Zelko et. al. have spearheaded most recent developments within OHS via the creation of a variety of tools and demonstrated their effectiveness in using these tools for novel observational health research endeavors \cite{zelko2022pilot}
Through such efforts, OHS is becoming further capable of executing novel observational health research.

% TODO: Cite FunOHDSI
% TODO: Cite OHDSICohortExpressions
% 
% TODO: Cite OHDSI Symposium poster on JuliaHealth
% TODO: Register OHDSI Symposium poster to Zenodo
% TODO: Cite JuliaCon 2023 presentation
% TODO: Register JuliaCon 2023 presentation to Zenodo

@@

>>>>

% EXPLAIN: Why is this a crucial innovation?
% EXPLAIN: What is meant by a transferable analysis
Arguably, the most crucial innovation that has emerged is the Observational Medical Outcomes Partnership Common Data Model (OMOP CDM) which was designed specifically to standardize observational health data to enable transferable analyses.
% TODO: Add citation
% NOTE: Why is mapping relevant here?
The importance of this model has become apparent within open science organizations like the OHDSI (Observational Health Data Science and Informatics) network where members across $100+$ countries with access to various health systems have been able to successfully map their patient data to the OMOP CDM.
% TODO: Add citation about successful collaborations
% EXPLAIN: What is a network study?
% TODO: Cite the quote "write once, run everywhere"
This has led to successful collaboration relationships across countries where one research group can develop a network study and then disseminate the study across collaborators leading to the idea that you can "write once, run everywhere."

\subsection{Background}

@@

@@

% TODO: Cite Discourse post on start of JuliaHealth
JuliaHealth is an open-science collective dedicated to bringing the Julia programming language to various avenues of health research domains including, but not limited to, observational health, medical imaging, standards interoperability and more.
It leverages the strengths of the Julia language to introduce new methods into these spaces, augment existing workflows, and provide a community structure for health researchers within the global Julia ecosystem.
% EXPLAIN: What is a subecosystem?
% EXPLAIN: How does having more subecosystems enable novel research to be done?
Eventually, JuliaHealth will host a variety of subecosystems and associated resources that will enable novel research to be conducted within Julia.

% TODO: Move this to What Is JuliaHealth section
As JuliaHealth is an ad hoc network of contributors, varying aspects of the ecosystem have developed at differing rates as a function of factors such as developer time or workplace incentives.

One area of the subecosystems within JuliaHealth that has matured significantly over the years has been the JuliaHealth observational health subecosystem (OHS).
It can be said that OHS had its inception in work done by Evans and Simonov with efforts such as \textit{FunOHDSI.jl}, \textit{OHDSICohortExpressions.jl}, and \textit{FunSQL.jl} around $2021$ \cite{kirill_simonov_2023_7705325}
% TODO: Cite FunOHDSI
% TODO: Cite OHDSICohortExpressions
% TODO: Cite OHDSI Symposium poster on JuliaHealth
% TODO: Register OHDSI Symposium poster to Zenodo
% TODO: Cite JuliaCon 2023 presentation
% TODO: Register JuliaCon 2023 presentation to Zenodo
Since then, Zelko et. al. have spearheaded most recent developments within OHS via the creation of a variety of tools and demonstrated their effectiveness in using these tools for novel observational health research endeavors \cite{zelko2022pilot}
% EXPLAIN: What does this mean?
Through such efforts, OHS is becoming further capable of executing novel observational health research.
